Program Transformations for miniKanren

### 2020-06-30 papers/aplas/evaluation.tex
Added ecce running time for unification and path search

\section{Evaluation}

In our study we compared the CPD adaptation for \mk{} and the new non-conjunctive partial deduction.
We have also employed the branching heuristic instead of the deterministic unfolding in the CPD to check whether it can improve the quality of the specialization.

We used the following 4 programs to test the specializers on.
\begin{itemize}
  \item Two implementations of an evaluator of logic formulas.
  \item A program to compute a unifier of two terms.
  \item A program to search for paths of a specific length in a graph.
\end{itemize}

The last two relations are described in~\cite{lozov2019relational} thus we will not describe them here.

@@

\subsection{Evaluation Results}

In our study, we considered two implementations of \lstinline{eval$^o$}, one we call \lstinline{plain} and the other --- \lstinline{last}, and compared how specializers behave on them.
The \lstinline{plain} relation uses table-based boolean operations and places them further to the left in each conjunction.
The relation \lstinline{last} employs boolean operations implemented via \lstinline{nand$^o$} and place them at the end of each conjunction.
These two programs are complete opposites from the standpoint of CPD.

We measured the time necessary to generate $10000$ formulas over two variables which evaluate to \lstinline{^true}.
We compared the results of specialization of the goal \lstinline{eval$^o$ subst fm ^true} by our implementation of CPD, the new non-conjunctive partial deduction, and the CPD modified with the branching heuristic.
Our evaluation confirmed that CPD behaves very differently on these two implementations of the same relation: the running time of the specialized \lstinline{last} relation is almost 6 times as high as the running time of the specialized \lstinline{plain} relation.
The running time of two programs generated with our novel non-conjunctive partial deduction is very close and it is a little bit better than the best by CPD.
CPD with the branching heuristic still constructs residual programs of different quality.
The results are shown in table~\ref{tbl:eval}.

@@

The \lstinline{unify} relation was executed to find a unifier of two terms $f(X, X, g(Z, t))$ and $f(g(p, L), Y, Y)$.
The original \mk{} program fails to terminate on this goal in 300 seconds.
On this example, the most performant is the program generated by CPD (2.35 seconds) while the program generated by adding branching heuristic also fails to terminate in 300 seconds.
The non-conjunctive partial deduction shows some improvement with the residual program terminated within 15 seconds.
While driving this program, the non-conjunctive partial deduction does too much unfolding which negatively impacts the running time as compared to CPD.

The last test executed \lstinline{isPath} relation to search for 5 paths in the graph with 20 vertices and 30 edges.
On this program, non-conjunctive partial deduction showed better transformation results than CPD, although the difference is not that drastic.

All evaluation results are presented in the table~\ref{tbl:eval}.
Each column corresponds to the relation being run as described above.
The row marked ``Original'' contains the running time of the original \mk{} relation before specialization, ``CPD'' and ``Non-CPD'' correspond to conjunctive and non-conjunctive partial deduction respectively while ``Branching'' is for the CPD modified with the branching heuristic.

\begin{table}
  \centering
  \begin{tabular}{c||c|c||c||c}
                   & last & plain & unify & isPath \\
  \hline\hline
  Original         & >60.00s & >60.00s & >300.00s & 19.86s \\
  \hline
  CPD              & 31.31s & 5.46s & 2.35s & 4.66s \\
  \hline
  Non-CPD          & 4.99s  & 5.05s & 14.90s & 3.00s \\
  \hline
  Branching        & 17.21s  & 6.17s & >300.00s & N/A \\
  \hline
  \end{tabular}

% \subsection{Search for a Unifier}

% % \begin{table}
% %   \centering
% %   \begin{tabular}{c|c}

% %     & running time \\
% %   \hline\hline
% %   Original         & >300.00s \\
% %   \hline
% %   CPD              & 2.35s \\
% %   \hline
% %   Non-CPD          & 14.90s \\
% %   \hline
% %   Branches         & >300.00s \\
% %   \hline
% %   \end{tabular}

% %   \caption{Searching for a unifier for terms f(X, X, g(Z, t)) and f(g(p, L), Y, Y)}
% %   \label{tbl:unify}
% % \end{table}

% The unification of two terms $t$ and $u$ is searching for a substitution $\theta$ such that $t \theta = u \theta$, $\theta$ is called a unifier.
% We search for any unifier, not necessarily most specific.
% The details on this benchmark can be found in the earlier paper CITE.
% This example demonstrates how too much unfolding can be introduced with the non-conjunctive partial deduction.

% \subsection{Search for Paths in a Graph}

% Here we search for 5 paths in a graph.
% The details on this benchmark can be found in the earlier paper CITE.

% % \begin{table}
% %   \centering
% %   \begin{tabular}{c|c}

% %     & running time \\
% %   \hline\hline
% %   Original         & 19.86s \\
% %   \hline
% %   CPD              & 4.66s \\
% %   \hline
% %   Non-CPD          & 3.00s \\
% %   \hline
% %   \end{tabular}

% %   \caption{Searching for paths in a graph}
% %   \label{tbl:unify}
% % \end{table}

>>>>

\section{Evaluation}
In our study we compared the CPD adaptation for \mk{} and the new non-conjunctive partial deduction.
We have also employed the branching heuristic instead of the deterministic unfolding in the CPD to check whether it can improve the quality of the specialization.

We used the following programs to test the specializers on.
\begin{itemize}
  \item Two implementations of an evaluator of logic formulas.
  \item Two implementations of a typechecker for a small expression language.
  \item A program to compute a unifier of two terms.
  \item A program to search for paths of a specific length in a graph.
\end{itemize}

The last two relations are described in~\cite{lozov2019relational} thus we will only briefly describe them here.

We focused on these particular examples because they are

\begin{itemize}
  \item Relational interpreters
  \item Natural
  \item Observable
\end{itemize}

@@

\subsubsection{Evaluation Results}
In our study, we considered two implementations of \lstinline{eval$^o$}, one we call \lstinline{plain} and the other --- \lstinline{last}, and compared how specializers behave on them.
The \lstinline{plain} relation uses table-based boolean operations and places them further to the left in each conjunction.
The relation \lstinline{last} employs boolean operations implemented via \lstinline{nand$^o$} and place them at the end of each conjunction.
These two programs are complete opposites from the standpoint of CPD.
We measured the time necessary to generate $1000$ formulas over two variables which evaluate to \lstinline{^true} (averaged over 10 runs).

\begin{table}
  \centering
  \begin{tabular}{c||c||c}
                   & last    & plain  \\
  \hline\hline
  Original         & 0.695s  & 1.426s \\
  \hline
  CPD              & 1.721s  & 0.168s \\
  \hline
  Ecce             & 0.205s  & 0.264s \\
  \hline
  Non-CPD          & 0.132s  & 0.137s \\
  \hline
  Branching        & 0.468s  & 0.179s \\
  \end{tabular}

\subsection{Unification and Path Search}

@@

The \lstinline{unify} relation was executed to find a unifier of two terms $f(X, X, g(Z, t))$ and $f(g(p, L), Y, Y)$.
The original \mk{} program fails to terminate on this goal in 30 seconds.
On this example, the most performant is the program generated by CPD (0.352 seconds) while the program generated by adding branching heuristic also fails to terminate in 30 seconds.
The non-conjunctive partial deduction shows some improvement with the residual program terminated within 1.947 seconds.
While driving this program, the non-conjunctive partial deduction does too much unfolding which negatively impacts the running time as compared to CPD.

The last test executed \lstinline{isPath} relation to search for 3 paths of length 7 in the graph with 20 vertices and 30 edges.
On this program, non-conjunctive partial deduction showed better transformation results than CPD, although the difference is not that drastic.

All evaluation results are presented in the table~\ref{tbl:unify}.
Each column corresponds to the relation being run as described above.
The row marked ``Original'' contains the running time of the original \mk{} relation before specialization, ``CPD'' and ``Non-CPD'' correspond to conjunctive and non-conjunctive partial deduction respectively while ``Branching'' is for the CPD modified with the branching heuristic.

\begin{table}
  \centering
  \begin{tabular}{c||c|c|c||c}
                   & \multicolumn{3}{c}{unify} & isPath \\
  \hline
                   & first  & second & third   &        \\
  \hline\hline
  Original         & $0.247*10^{-4}s$ & $>30s$ (oum?) & $>30s$ (oum?) & $23.743s$ \\
  \hline
  CPD              & $0.081*10^{-4}s$ & $0.017s$ & $0.343s$  & $2.999s$  \\
  \hline
  Non-CPD          & $0.084*10^{-4}s$ & $0.021s$ & $2.198s$  & $1.786s$  \\
  \hline
  Ecce             & $0.128*10^{-4}s$ & $0.026s$ & $1.340s$  & $1.292s$ \\
  \hline
  Branching        & $0.205*10^{-4}s$ & $0.101s$ & $>30s$ (oum?) & N/A \\
  \hline
  \end{tabular}

  \caption{Evaluation results of unification and path search}
  \label{tbl:unify}
\end{table}

\subsection{Typechecker-Term Generator}

This relation implements a typechecker for a tiny language. Being executed in the backward direction it serves as a generator of terms of the given type.

Its abstract syntax is the following.

% \begin{align*}
%   &type \ term = \\
%   &| \ BConst \ of \ Bool \\
%   &| \ IConst \ of \ Int \\
%   &| \ Var \ of \ Int \\
%   &| \ Plus \ of \ term * term \\
%   &| \ Mult \ of \ term * term \\
%   &| \ Eq \ of \ term * term \\
%   &| \ Lt \ of \ term * term \\
%   &| \ Let \ of \ term * term \\
%   &| \ If \ of \ term * term * term
% \end{align*}

\begin{align*}
  &type \ term = \\
  &| \ BConst \ of \ Bool \\
  &| \ IConst \ of \ Int \\
  &| \ Var \ of \ Int \\
  &| \ term + term \\
  &| \ term * term \\
  &| \ term = term \\
  &| \ term < term \\
  &| \ \underline{let} \ term \ \underline{in} \ term \\
  &| \ \underline{if} \ term \ \underline{then} \ term \ \underline{else} \ term
\end{align*}

We use de Bruijn indices for variables.

Typing rules are the following

\begin{table}
  \setlength{\tabcolsep}{0.5cm}
  \centering
  \begin{tabular}{c c}
    \infer[]{\Gamma \vdash IConst \ i : Int}{} &
    \infer[]{\Gamma \vdash BConst \ b : Bool}{} \vspace{0.5cm} \\

    \infer[]{\Gamma \vdash t + s : Int}{\Gamma \vdash t : Int, \Gamma \vdash  s : Int} &
    \infer[]{\Gamma \vdash t * s : Int}{\Gamma \vdash t : Int, \Gamma \vdash  s : Int} \vspace{0.5cm} \\

    \infer[]{\Gamma \vdash t = s : Bool}{\Gamma \vdash t : \tau, \Gamma \vdash  s : \tau} &
    \infer[]{\Gamma \vdash t < s : Bool}{\Gamma \vdash t : Int, \Gamma \vdash  s : Int} \vspace{0.5cm} \\

    \infer[]{\Gamma \vdash \underline{let} \ v \ b : \tau}{\Gamma \vdash v : \tau_v, \ (\tau_v :: \Gamma) \vdash b : \tau} &
    \infer[\Gamma \lbrack v \rbrack = \tau]{\Gamma \vdash Var \ v : \tau}{} \vspace{0.5cm} \\

    \multicolumn{2}{c}{
      \infer[]{\Gamma \vdash \underline{if} \ c \ \underline{then} \ t \ \underline{else} \ s : \tau}{\Gamma \vdash c : Bool, \Gamma \vdash t : \tau, \Gamma \vdash s : \tau}
    }

  \end{tabular}
\end{table}

Implementation 1 is a straightforward implementation in a relational programming language.

Implementation 2 is obtained by translating into \mk{} an implementation of the typing rules \ocaml{}.

We measured the time needed to generate 100 formulas (averaged over 10 iterations).

\begin{table}
  \centering
  \begin{tabular}{c||c||c}
              & Implementation 1 & Implementation 2 \\
  \hline\hline
  Original    & 0.057s & 0.464s\\
  \hline
  Non-CPD     & 0.154s & 0.448s \\
  \hline
  Ecce        & 2.543s  & 1.518s \\
  \end{tabular}

  \caption{Running time of generating 100 closed terms of type Int}
  \label{tbl:eval}
\end{table}
